Estimation of the Effect of a Continuous Treatment Variable

### 2025-11-18 Writing/manuscript_CF_continuous.qmd
Working on comments

@@

$$
\begin{align*}
y = f(X, N) + g(W) + v
\end{align*}
$$

@@

In the single-orthogonalized framework only the baseline component $g(W)$ is partialed out before the final-stage estimation; the continuous treatment $N$ and heterogeneity-inducing covariates $X$ remain in their original form. Formally, starting from Model 3, we first estimate $g(W)=\mathbb{E}[y \mid W]$ using a cross-fitted random forest and form the residualized outcome

$$
\begin{align*}
\tilde{y}=y-\widehat{g}(W)
\end{align*}
$$

@@

To adapt double-orthogonalization to the model of interest (Model 3), we estimate the heterogeneous response $f(X,N)$ by expanding the continuous treatment $N$ in a set of spline basis functions and allowing their coefficients to vary with the heterogeneity-inducing covariates $X$:

$$
\begin{align*}
y=\sum_{k=1}^K \theta_k(X) S_k(N)+g(X, W)+v
\end{align*}
$$

Although double-orthogonalization generally fails when $f(X,N)$ is an arbitrary nonlinear function of $N$, this spline-based formulation restores the key linear-in-parameter structure that makes DO valid, because the model is linear in the transformed treatment basis $\left\{S_k(N)\right\}$. Residualizing each basis component and the outcome with respect to $(X,W)$ then yields a clean decomposition analogous to the linear-in-$T$ case. Let $\widetilde{S}_k(N)=S_k(N)-\mathbb{E}\left[S_k(N) \mid X, W\right]$. Then

$$
\begin{align*}
\tilde{y}=y-\mathbb{E}[y \mid X, W]=\sum_{k=1}^K \theta_k(X) \widetilde{S}_k(N)+v,
\end{align*}
$$

so causal forest can nonparametrically estimate each heterogeneous weight $\theta_k(X)$  by relating the residualized outcome to the residualized spline bases. This Neyman-orthogonal (R-learner) style decomposition preserves the ability to capture complex, heterogeneous, and nonlinear treatment effects while inheriting the robustness of double-orthogonalization: small errors in the nuisance fits for $g(X,W)$ or the conditional means of the bases do not induce first-order bias in $\theta_k(X)$. The estimated $\theta_k(X)$ then combine to recover the full heterogeneous response function and, by differencing across candidate $N$ values, site-specific treatment effects and economically optimal nitrogen rates.

>>>>

new

On-farm precision experiments (OFPE) have emerged as a powerful approach to generate the data necessary for site-specific nitrogen management (@bullock2019data; @lacoste2022farm). Through on-farm agronomic trials, researchers can directly observe yield responses to varying nitrogen rates under real-world field conditions, enabling estimation of yield response functions and the identification of economically optimal nitrogen rates. In recent years, prediction-oriented machine learning (ML) methods have been widely adopted in this context to model yield as a function of nitrogen and environmental covariates (@krause2020random; @barbosa2020modeling; @barbosa2020risk; @gardner2021economic). However, recent evidence suggests that prediction-oriented ML approaches, despite often achieving high yield-prediction accuracy, do not necessarily recover the underlying yield response to nitrogen. @tanaka2024can show that fertilizer rate recommendations derived from predictive ML models are highly sensitive to algorithm choice and covariate selection, leading to substantial variability in estimated optimal rates. These findings indicate a critical difference between predicting yield levels and estimating the causal yield response that determines economically optimal nitrogen rates, motivating the use of methods that explicitly estimate causal effects of nitrogen on yield.

@@

$$
\begin{align*}
y = f(X, N) + g(X,W) + v
\end{align*}
$$

@@

In the single-orthogonalized framework only the baseline component $g(X,W)$ is partialed out before the final-stage estimation; the continuous treatment $N$ and heterogeneity-inducing covariates $X$ remain in their original form. Formally, starting from Model 3, we first estimate $g(X,W)=\mathbb{E}[y \mid X,W]$ using a cross-fitted random forest and form the residualized outcome

$$
\begin{align*}
\tilde{y}=y-\widehat{g}(X,W)
\end{align*}
$$

@@

To adapt double-orthogonalization to the model of interest (Model 3), we estimate the heterogeneous response $f(X,N)$ by expanding the continuous treatment $N$ in a set of spline basis functions and allowing their coefficients to vary with the heterogeneity-inducing covariates $X$.

In spline regression, a single variable can be decomposed into multiple basis functions in a linear-in-parameter manner. For example consider the following general model, where the non-linear impact of $T$ is represented by $f(T)$.

$$
y = f(T) + v 
$$

Non-linearity in $f(T)$ can be captured by a linear combination of basis functions as follows:

$$
f(T) = \alpha_1 \cdot S_1(T) + \alpha_2 \cdot S_2(T) \dots, + \alpha_K \cdot S_K(T) = \sum_{k=1}^K \alpha_k \cdot S_k(T)
$$

where $\alpha_k$ is the coefficient to be estimated and $S_k(T)$ is the value of the basis function $S_k(\cdot)$ evaluated at $T$. By changing the value of $\alpha_k$, $\sum_{k=1}^K \alpha_k S_k(T)$ can flexibly approximate various non-linear functions that cannot be captured by commonly used functional forms. In spline regression, the weights (or coefficients) are obtained by regression with the spline bases included as covariates.

```{r fig-spline-examples, fig.cap="An example of spline regression fit of the yield response curved and weighted basis functions"}
knitr::include_graphics(here::here("writing/Figures/g_tsp_fit.pdf"))
```

@fig-spline-examples shows the fitted line (red line), which is the sum of the six weighted (multiplied by coefficients) thin-plate splines (dotted lines). As can be seen, the fitted line closely traces the the true yield response function (blue line).

Since this modeling approach is linear-in-parameter, it can be readily used inside the CF framework. Instead of a constant as the weight (coefficient) for each spline basis function, the weight is made a function of observables, $X$, as follows:

$$
\begin{align*}
y=\sum_{k=1}^K f_k(X) S_k(N)+g(X, W)+v
\end{align*}
$$

Although double-orthogonalization generally fails when $f(X,N)$ is an arbitrary nonlinear function of $N$, this spline-based formulation restores the key linear-in-parameter structure that makes DO valid, because the model is linear in the transformed treatment basis $\left\{S_k(N)\right\}$. Residualizing each basis component and the outcome with respect to $(X,W)$ then yields a clean decomposition analogous to the linear-in-$T$ case.

Let $\widetilde{S}_k(N)=S_k(N)-\mathbb{E}\left[S_k(N) \mid X, W\right]$. Then

$$
\begin{align*}
\tilde{y}=y-\mathbb{E}[y \mid X, W]=\sum_{k=1}^K f_k(X) \widetilde{S}_k(N)+v,
\end{align*}
$$

so causal forest can nonparametrically estimate each heterogeneous weight $f_k(X)$  by relating the residualized outcome to the residualized spline bases. This Neyman-orthogonal (R-learner) style decomposition preserves the ability to capture complex, heterogeneous, and nonlinear treatment effects while inheriting the robustness of double-orthogonalization: small errors in the nuisance fits for $g(X,W)$ or the conditional means of the bases do not induce first-order bias in $f_k(X)$. The estimated $f_k(X)$ then combine to recover the full heterogeneous response function and, by differencing across candidate $N$ values, site-specific treatment effects and economically optimal nitrogen rates.
